Left-linear rules in adhesive categories

### 2024-02-13 icalp/fusione.tex
lemmi 3 passi

  A derivation $\der{D}$ consisting of three steps $\dder{D}_0$,
  $\dder{D}_1$, and $\dder{D}_2$, each applying the corresponding rule
  $\rho_i$, is depicted in Fig.~\ref{fi:derD}. The second and third
  steps $\dder{D}_1$ and $\dder{D}_2$ are clearly sequential
  independent, via an independence pairs that map nodes consistently
  with their numbering.\todo{P: having names for the graphs would
    probably help}

@@

As already mentioned, for linear rewriting systems, when viewing
derivations as concurrent computations, it is canonical to identify
derivations which are equal ``up to switching'', i.e., which differs
only for the order of indepedent steps. The same notion can be given
for left-linear rewriting systems by relying on notion of switch.

@@

Switch equivalence is now defined as the equivalence relating derivations which can be obatained one from another by a sequence of switches. Moreover, intermediate graphs can be taken up to isomorphism according to abstraction equivalence.

@@

\begin{example}
  \label{ex:seq-ind}
  Consider a rewriting system in $\cat{Graph}$, the category of
  directed graphs and graph morphisms, which is adhesive. The set of
  rules $\R = \{ \rho_0, \rho_1, \rho_2\}$ consists of three rules as
  depicted in Fig.~\ref{fi:rules}, where numbers are used to represent
  the mappings from the interface to the left- and right-hand
  sides. Rules $\rho_0$ and $\rho_1$ are linear: $\rho_0$ generates a
  new node, while $\rho_1$ deletes an edge. Instead, $\rho_2$ is non
  linear as it ``merges'' two nodes.

@@

  A derivation $\der{D}$ consisting of three steps $\der{D}_0$,
  $\der{D}_1$, and $\der{D}_2$, each applying the corresponding rule
  $\rho_i$ is depicted in in Fig.~\ref{fi:derD}. The second and third
  steps $\der{D}_1$ and $\der{D}_2$ are clearly sequential
  independent, via an independence pair $L_2 \to D_1$ and
  $R_1 \to G_2$ that map nodes accoording to their numbering.

@@

%\rem{Non introdurrei una classe, come detto sopra}{
\begin{definition}[Strong Enforcing rewriting systems]A left-linear DPO rewriting system is \emph{strong enforcing} if every independence pair between two direct derivations is strong. }
\end{definition}

@@

As a sanity check note that a source of strong enforcing left-linear DPO rewriting systems is given by the linear ones, as shown by the following proposition.

@@

Even if we work in strong enforcing rewriting systems where sequential
independence ensures switchability, when dealing with left-linear
rules there is a further -- and possibly more serious -- issue, i.e.,
the fact that there can be more than one independence pairs between
the same derivations. This can hinder the core idea of using
sequential independence as a basis of a theory of concurrency for
rewriting systems, since exchanges performed using different
independence pairs may lead to derivations that are not abstraction
equivalent, and thus one would equate computations that should
definitively be taken apart, as shown in the example below.

\begin{example}
  Consider the derivation $\der{E}$ from Example~\ref{ex:seq-ind}.
  %
  The last two steps are sequential independent but one easily sees
  that there are two distinct indepedence pair, as the left-hand side
  of $\rho_1$ can be mapped either to node $1$ or to node
  $2$ in $G_2'$. Correspondingly, there
  are two switches of $\der{E}$: one is the derivation $\der{D}$ in
  Figure~\ref{fi:derD} we started from, the other is the derivation
  $\der{D}'$ in Figure~\ref{fi:derD1}.

@@

  As a consequence, $\der{D}$ and $\der{D}'$ are switch equivalent, but
  this is not acceptable when viewing equivalence classes of
  derivations as concurrent computations: in $\der{D}$ the first two
  steps are not sequential independent, while in $\der{D}'$ they are,
  intuitively because in $\der{D}$ rule $\rho_1$ uses the node
  generated by $\rho_0$, while in $\der{D}'$ rule $\rho_1$ uses the
  node which was in the initial graph. From the technical point of
  view, the property of being switch equivalent is not closed by
  prefix, something which prevents to derive a sensible concurrent
  sematics: if we limit to the first two steps, derivations
  $\dder{D}_0; \dder{D}_1$ and $\dder{D}_0'; \dder{D}_1'$ are not switch
  equivalent while $\der{D}$ and $\der{D}'$ would be
\end{example}

\full{Moreover, limiting sequential independence to the case in which the independence
pair is unique (as suggested in~\cite{baldan2017domains}) brings to the same problems.}

@@

The previous definition might sound restrictive. However, it is immediate to see that linear rewriting systems are all well switching (see Proposition~\ref{pr:weak}).
%
Moreover, we next introduce a class of rewriting systems, comprising
all the ones used in concrete modelling applications of DPO rewriting
(e.g., in the encoding of process calculi ..., ...) are \todo{INSERIRE
  ESEMPI QUI} actually well switching.
%
Roughly speakings, these are possibly hierarchical graphical structures where rules are contrained not to merge elements of top level sorts in the hiearchy (for graphs, nodes can be merged while edges can not).

@@

The next technical lemma is at the core of globality for independence. It shows that if we have a derivation consisting of three
rewriting steps, say $r_0\, r_1\, r_2$ where $r_0$ and $r_1$ are independent, then if we switch $r_2$ to the first position, 
obtaining a derivation $r_2\, r_0\, r_1$, the steps $r_0$ and $r_1$ are still independent.

Instead, if in $r_0\, r_1\, r_2$ we have $r_1$ and $r_2$ are
independent, and we switch $r_0$ to the last position, obtaining a
derivation $r_1\, r_2\, r_0$, the steps $r_1$ and $r_2$ could be no
longer independent, unless an additional condition is
satisfied. Intuitively this is due to the fact that, dealing with
left-linear rewriting systems introduces disjunctive forms of
causality. Hence, if in $r_0\, r_1\, r_2$ we have that $r_0, r_1$ are
disjunctive causes of $r_2$, at least one of the two is needed to
enable the application of $r_2$ and this explain why in
$r_1\, r_2\, r_0$ we have that $r_1$ and $r_2$ are no longer
independent.

@@

\begin{restatable}[Globality of independence]{lemma}{lemIndepGlobalLeft}
  \label{lem:indep-global-left}
  Let $\X$ be an $\mathcal{M}$-adhesive category and
  $(\X,\R)$ a left-linear DPO rewriting system.
  Consider a derivation
  $\der{D}=\{\dder{D}_i\}_{i=0}^2$ and suppose that $(i_0,i_1)$ is a
  strong independence pair between $\dder{D}_0$ and $\dder{D}_1$,
  $(a_0,a_1)$ one between $\dder{D}_1$ and $\dder{D}_2$ and
  $(e_0, e_1)$ one between $\dder{D}_0$ and
  $S_{a_0,a_1}(\dder{D}_2)$. Then the following properties hold true.
  \begin{enumerate}
  \item
    \label{lem:indep-global-left:1}
    $S_{e_0,e_1}(\dder{D}_0)$ and $S_{a_0,a_1}(\dder{D}_1)$ are
    sequentially independent.

  \item
    \label{lem:indep-global-left:2}
    If $S_{i_0, i_1}(\dder{D}_0)\seqind \dder{D}_2$ with a
    strong independence pair $(\alpha_0, \alpha_1)$, then
    $S_{i_0,i_1}(\dder{D}_1)$ and $S_{\alpha_0, \alpha_1}(\dder{D}_2)$
    are  sequentially independent.
  \end{enumerate}
\end{restatable}

\todo{P: anche qui sopra mi limiterei a well switching, cos\i' da evitare riferimenti a strong e se si riesce anche alle indepdencen pairs}
  %

@@

\section{Conclusions and further work}

We perform an in-depth investigation of notion of independence between
computational steps in the setting of left-linear rewriting systems
over adhesive categories, which encompasses rewriting over most of the
structures used in semantic and modelling applications.

We showed that the canonical notion of independence adopted for linear
systems does not enjoy some properties which are essential for
developing a sensible theory of concurrency, notably Church Rosser
theorem, and we identified the subclass of well switchable rewriting
systems as an appropriate setting where instead a number of vital
results can be re-established. Specifically, Church Rosser theorem,
i.e., the possibility of performing independent steps in any order is
fully recovered. Moreover, the switch construction and correspondingly
switch equivalence which is at the core of a theory of concurrency for
adhesive recwritng suystems enjoy a number of fundamental properties:
confluence of switching, a weak form of globality of independence and
the existence of a canonical form for switch equivalence proofs.

The class of well-switchabke adhesive rewting systems is vast. Generalising a result in~\cite{baldan2011adhesivity} we can show that one of its defining properties (the fact that sequential independence implies switchability) holds for a class of adhesive categories which include all presheaves over set. Moreover, the second property, uniqueness of switching, is ensured when rewriting possibly hierarchical graphical structures, as long as elements of the top sorts are not merged (e.g., one can consider rewriting systems over directed graphs where only nodes are merged, as it happens in most approaches to modelling calculi, biological systems and also in string diagrams).\todo{Giusto, referneces?}

The results in this paper open the way to the development of a
concurrent semantics for left-linear rewriting systems over adhesive
categories. In~\cite{baldan2017domains} the authors argue that for
computational systems which allow one to express merging or fusions,
the right semantic models are weak prime domains, a generalisation of
of prime domains, a staple in concurrency theory, and connected event
structures, their event-based counterpart. The mentioned paper
discusses only the case of graph rewriting and does not delve into the
problems induced by merging to the trheory of rewriting (the systems
considered there are implicitly assumed to be well switching). We plan to consolidate the statement that weak prime domains are ``the'' model for systems with fusions by showing that they allow to provide a semantics for general left-linear rewriting systems in $\mathcal{M}$-adhesive categioris.

On the practical side, the present paper just surveyed the possibility
of modelling Egg graphs. This appears to be quite interesying as
concurrent rewriting on Eggs seems to be an actove area of
reserach. \todo{P: cite} We plan to make the correspondence precise
and investigate how the analysis techniques enabled by a concurrent
semantics of rewriting can impact on Eggs.

@@

\noindent \parbox{7.4cm}{
\begin{lemma}\label{lem:varie}Let $\mathcal{A}$ be a class of arrows stable under pullbacks, pushouts and containing all isomorphisms.  Suppose that, the left square aside is $\mathcal{A}$-Van Kampen, while the vertical faces in the right cube are pullbacks. 	Supppose, moreover that $m\colon A\to C$ and $d\colon D'\to D$ are mono and that $d$ belongs to $\mathcal{A}$. Then $d\leq n$ if and only if $c \leq m$.
\end{lemma}}
	\parbox{6cm}{$\xymatrix@C=10pt@R=10pt{&&&&&A'\ar[dd]|\hole_(.65){a}\ar[rr]^{g'} \ar[dl]_{m'} && B' \ar[dd]^{b} \ar[dl]_{n'} \\ A \ar[dd]_{m}\ar[rr]^{g}&&B\ar[dd]^{n}&&C'  \ar[dd]_{c}\ar[rr]^(.7){f'} & & D' \ar[dd]_(.3){d}\\&&&&&A\ar[rr]|\hole^(.65){g} \ar[dl]^{m} && B \ar[dl]^{n} \\C\ar[rr]_{f} &&D&&C \ar[rr]_{f} & & D}$}

@@

	\smallskip \noindent
	$(\Leftarrow)$ Let $h:C'\to A$ be such that $c=m\circ h$. By the $\mathcal{A}$-Van Kampen property the top face of the given cube is a pushout. Thus the dotted $k:D'\to B$ in the following diagram exists.
	\[\xymatrix@C=40pt{A' \ar@/_.5cm/[dd]_{a} \ar[d]^{m'} \ar[r]^{g'} & B' \ar[d]_{n'} \ar@/^.5cm/[dd]^{b} \\  C' \ar[d]^{h} \ar[r]_{f'} & D' \ar@{.>}[d]_{k}\\ A \ar[r]_g& B }\]
	Moreover, by construction we have
	\begin{align*}
		n\circ k \circ n' & = n\circ b=d\circ n'                                  \\
		n\circ k \circ f' & = n\circ g\circ h=f\circ m\circ h=f\circ c= d\circ f'
	\end{align*}
	We can therefore conclude that $n\circ k =d$.
\end{proof}}

@@

\full{ 
\begin{proof}
	We can begin noticing that $g$, being the pullback of $k$, is mono and in $\mathcal{A}$ too. Thus we can build the cube below, in which all the vertical faces are pullbacks, entailing that all the vertical arrows are in $\mathcal{A}$.
	\[\xymatrix@C=20pt@R=10pt{ & &X\ar[dl]_{f} \ar[ddd]^(.333333){\id{X}}|(.666666)\hole\ar[rrr]^{a} &&&A \ar[ddd]^{\id{A}} \ar[dl]^{h}\\& Y\ar[dl]_{\id{Y}} \ar[ddd]|(.333333)\hole_{\id{Y}} &&&B \ar[ddd]^{\id{B}} \ar[dl]_{\id{B}} \\Y \ar[ddd]_{g} \ar[rrr]^{b}&&& B\ar[ddd]^{k}\\&&X \ar[rrr]|(.34)\hole^{a}|(.67)\hole \ar[dl]_{f}&&& A \ar[dl]^{h}\\ & Y  \ar[dl]_{g}&&& B \ar[dl]^{k}\\ Z\ar[rrr]_{c} &&& C}\]
	By hypothesis the bottom face is an $\mathcal{A}$-stable pushout and so its top one is a pushout. Using \Cref{lem:po1} we can conclude that the right half of the rectangle we  started with is a pushout too.
\end{proof}}

@@

\begin{corollary}\label{prop:bal}
	If $\X$ is an $\mathcal{M}$-adhesive categories, then every epimorphism in $\mathcal{M}$ is an isomorphism. In particular, every adhesive category $\X$ is \emph{balanced}: if a morphism is monic and epic, then it is an isomorphism.
\end{corollary}

@@

\noindent
\parbox{11cm}{
	\begin{definition}[Pushout complement]
		Let $f\colon X\to Y$ and $g\colon Y\to Z$ be two composable arrows in a category $\X$. A \emph{pushout complement} for the pair $(f,g)$ is a pair $(h,k)$ with $h\colon X\to W$ and $k\colon W\to Z$ such that the square aside commutes and it is a pushout.
	\end{definition}} \quad
\parbox{4cm}{\vspace{.1cm}
	$\xymatrix{X \ar[r]^{f} \ar[d]_{h}& Y \ar[d]^{g} \\ W \ar[r]_{k}& Z}$}\\

@@

In $\mathcal{M}$-adhesive categories the pushout complement is unique up to iso, if it exists. 
% amend the second defect.\todo{A: }

@@

	Thus there exists a unique $t\colon A\to P$ as in the diagram
	\[\xymatrix{A \ar@{>->}[d]_{a} \ar@{.>}[dr]^{t}\ar[r]^{r} &B \ar@{>->}@/^.2cm/[dr]^{s} \\ M \ar@{>->}@/_.2cm/[dr]_{m}&P\ar[r]^{p_1}  \ar@{>->}[d]_{p_2}& N \ar@{>->}[d]^{n}\\ &X \ar[r]_{f} & Y}\]
	We can then consider the cube below, in which the front, right and back faces are pullbacks.
	\[\xymatrix@C=15pt@R=15pt{&A\ar@{>->}[dd]|\hole_(.65){a}\ar[rr]^{r} \ar[dl]_{t} && B \ar@{>->}[dd]^{b} \ar[dl]_{s} \\ P  \ar@{>->}[dd]_{p_2}\ar[rr]^(.7){p_1} & & N \ar@{>->}[dd]_(.3){n}\\&M\ar[rr]|\hole^(.65){h} \ar@{>->}[dl]^{m} && Q \ar@{>->}[dl]^{k} \\X \ar[rr]_{f} & & Y}\]
	We can notice that the left square is a pullback too. To see this it is enough to apply \Cref{lem:pb1} to the diagram
	\[\xymatrix{A  \ar@/^.4cm/[rr]^{s\circ r}\ar@{>->}[d]_{a} \ar[r]_{t}& \ar[r]_{p_1} P \ar@{>->}[d]^{p_2}& N \ar[d]^{n}\\ M \ar@/_.4cm/[rr]_{k\circ h}\ar@{>->}[r]^{m}& X \ar[r]^{f}& Y}\]

@@

\full{ 
\begin{proof}
	Consider the two pushout squares
	\[\xymatrix{X \ar[r]^{h_1}\ar@{>->}[d]_{m}& W_1 \ar@{>->}[d]^{k_1} & X \ar@{>->}[d]_{m} \ar[r]^{h_2}& W_2 \ar@{>->}[d]^{k_2}\\Y\ar[r]_{f}&Z &Y \ar[r]_{f}&Z}\]
	Since $m$ is in $\mathcal{M}$, \cref{prop:pbpoad} guarantees that they are both pullbacks.
	Since $m\leq m$, by \Cref{lem:radj} we get that now entails that $k_1\leq k_2$ and $k_2\leq k_1$. Thus there exists an isomorphism $\phi\colon W_1\to W_2$ such that $k_1=k_2\circ \phi$. To see that $h_2=\phi\circ h_1$, we can compute:
	\[
		k_2\circ \phi \circ h_1  = k_1\circ h_1= f\circ m= k_2\circ h_2\]

@@

\label{propUnique-proof}
\begin{proof}
  Both pairs $(k, f)$ and $(k', f')$ are pushout complements of $l$
  and $m$, thus, \Cref{lem:pocomp} yields an isomorphism
  $\phi_D\colon D\to D'$. Computing we have
  \[
    g'\circ \phi_D \circ k= g' \circ k'=h'\circ r
  \]

@@

There is another
useful properties of weakly sequentially independent direct
derivations which is worth mentioning.\todo{A: va tolto weakly?}

\noindent
\parbox{9.5cm}
{\begin{proposition}\label{lem:cose}Let
    $\der{D}=\{\dder{D}_i\}_{i=0}^1$ be a derivation, and suppose
    that $(i_0, i_1)$ is an independence pair between $\dder{D}_0$
    and $\der{D}_1$.  If $\der{E}=\{\dder{E}_i\}_{i=0}^1$ is a
    switch of $\der{D}$ along $(i_0, i_1)$, then there exists
    $q_0\colon P\to D_{\der{E},0}$ in the diagram
    aside \end{proposition}}
\parbox{4cm}{$\xymatrix{D_{\der{E},0} \ar@{>->}[d]_{f_{\der{E},0}}&P
    \ar@{.>}[l]_{q_0} \ar[r]^{p_1} \ar@{>->}[d]_{p_0}& D_1
    \ar@{>->}[d]^{f_1}\\ G_{0}&D_0 \ar@{>->}[l]^{f_0}\ar[r]_{g_0}&
    G_1}$}

\begin{proof}
  Consider the arrow $u_1\colon K_1\to P$ obtained using
  \Cref{prop:tec}. It fits in the diagram:
  \[\xymatrix{K_1 \ar[r]^{u_1} \ar@{>->}[d]_{l_1}&P
      \ar[r]^{\id{P}}\ar@{>->}[d]_{p_0}& P\ar@{>->}[d]^{p_0}\\L_1
      \ar[d]_{\id{L_1}}\ar[r]^{i_1}&D_0 \ar[r]^{\id{D_0}}
      \ar[d]_{\id{D_0}}& D_{0} \ar@{>->}[d]^{f_0}\\
      L_1\ar[r]^{i_1} \ar@/_.4cm/[rr]_{m_{\der{E},0}}&D_0
      \ar@{>->}[r]^{f_0}&G_0}\]

>>>>

  A derivation $\der{D}$ consisting of three steps $\dder{D}_0$,
  $\dder{D}_1$, and $\dder{D}_2$, each applying the corresponding rule
  $\rho_i$, is depicted in in Fig.~\ref{fi:derD}. The second and third
  steps $\dder{D}_1$ and $\dder{D}_2$ are clearly sequential
  independent, via an independence pairs that map nodes consistently
  with their numbering.\todo{P: having names for the graphs would
    probably help}

@@

As already mentioned, for linear rewriting systems, when viewing
derivations as concurrent computations, it is canonical to identify
derivations which are equal ``up to switching''. The same notion can
be given for left-linear rewriting systems by relying on notion of
switch.

@@

Switch equivalence is now defined as the equivalence relating derivations which can be obatained one from another by a sequence of switches. Moreover, intermediate graphs can be taken up to isomorphism.\todo{P: da scrivere meglio e fare}

@@

\begin{example}
  \label{ex:seq-ind}
  Consider a rewriting system in $\cat{Graph}$, the category of
  directed graphs and graph morphisms, which is adhesive. The set of
  rules $\R = \{ \rho_0, \rho_1, \rho_2\}$ consists of three rules as
  depicted in Fig.~\ref{fi:rules}, where numbers are used to represent
  the mappings from the interface to the left- and right-hand
  sides. Rules $\rho_0$ and $\rho_1$ are linear: $\rho_0$ generates a
  new node, while $\rho_1$ delete an edge. Instead, $\rho_2$ is non
  linear as it ``merge'' two nodes.

@@

  A derivation $\der{D}$ consisting of three steps $\der{D}_0$,
  $\der{D}_1$, and $\der{D}_2$, each applying the corresponding rule
  $\rho_i$ ,is depicted in in Fig.~\ref{fi:derD}. The second and third
  steps $\der{D}_1$ and $\der{D}_2$ are clearly sequential
  independent, via an independence pairs that map nodes consistently
  with their numbering.\todo{P: having names for the graphs would
    probably help}

@@

%\rem{Non introdurrei una classe, come detto sopra}{
\begin{definition}[Strong Enforcing rewriting systems]A left-linear DPO rewriting system is \emph{strong enforcing} if every independence pair between two direct derivations is strong.
\end{definition}

@@

As a sanity check not that a source of strong enforcing left-linear DPO rewriting systems is given by the linear ones, as shown by the following proposition.

@@

Even if we work in a setting where sequential independence ensures switchability, when dealing with left-linear rules there is a further -- and possibly more serious -- issue, i.e., the fact that there can be more than one independence pairs between the same derivations. This can hinder the core idea of using sequential independence as a basis of a theory of concurrency for rewriting systems, since exchanges performed using different independence pairs may lead to derivations that are not abstraction equivalent, and thus one would equate computations that 
should definitively be taken apart, as shown in the example below.

\begin{example}
  Consider the derivation $\der{E}$ from Example~\ref{ex:seq-ind}.
  %
  The last two steps are sequential independent but one easily sees
  that there are two distinct indepedence pair, as the left-hand side
  of $\rho_1$ can be mapped either to node $1$ or to node
  $2$.\todo{Sarebbe bello dire di quale grafo} Correspondingly, there
  are two switches of $\der{E}$: one is the derivation $\der{D}$ in
  Figure~\ref{fi:derD} we started from, the other is the derivation
  $\der{D}'$ in Figure~\ref{fi:derD1}.

@@

  As a consequence, $\der{D}$ and $\der{D}'$ are switch equivalent, but
  this is not acceptable when viewing equivalence classes of
  derivations as concurrent computations: in $\der{D}$ the first two
  steps are not sequential independent, while in $\der{D}'$ they are,
  intuitively because in $\der{D}$ rule $\rho_1$ uses the node
  generated by $\rho_0$, while in $\der{D}'$ rule $\rho_1$ uses the
  node which was in the initial graph. From the technical point of
  view, the property of being switch equivalent is not closed by
  prefix, something which prevents to derive a sensible concurrent
  sematics: if we limit to the first two steps, derivations
  $\der{D}_0; \der{D}_1$ and $\der{D}_0'; \der{D}_1'$ are not switch
  equivalent while $\der{D}$ and $\der{D}'$ would be
\end{example}

Moreover, limiting sequential independence to the case in which the independence
pair is unique (as suggested in~ ...) brings to the same problems.

@@

The previous definition might sound restrictive. However, it is immediate to see that linear rewriting systems are all well switching (see Proposition~\ref{pr:weak}).
%
Moreover, we next introduce a class of rewriting systems, comprising
all the ones used in concrete modelling applications of DPO rewriting
(e.g., in the encoding of process calculi ..., ...) are \todo{INSERIRE ESEMPI QUI}
actually well switching.

@@

The first technical lemma is at the core of globality for independence. It shows that if we have a derivation consisting of three rewriting steps, say $\dder{D}_0\cdot \dder{D}_1 \cdot \dder{D}_2$ where $\dder{D}_0$ and $\dder{D}_1$ are independent, then if we switch $\dder{D}_2$ to the first position, 
obtaining a derivation $\dder{D}_2 \cdot  \dder{D}_0\cdot\dder{D}_1$, the steps $\dder{D}_0$ and $\dder{D}_1$ are still independent.

Instead, if in $\dder{D}_0\cdot \dder{D}_1 \cdot \dder{D}_2$ we have $\dder{D}_1$ and $\dder{D}_2$ are
independent, and we switch $\dder{D}_0$ to the last position, obtaining a
derivation $\dder{D}_1\cdot \dder{D}_2\cdot  \dder{D}_0$, the steps $\dder{D}_1$ and $\dder{D}_2$ could be no
longer independent, unless an additional condition is
satisfied. Intuitively this is due to the fact that, dealing with
left-linear rewriting systems introduces disjunctive forms of
causality. Hence, if in $\dder{D}_0\cdot \dder{D}_1\cdot \dder{D}_2$ we have that $\dder{D}_0$ and $\dder{D}_1$ are
disjunctive causes of $\dder{D}_2$, at least one of the two is needed to
enable the application of $\dder{D}_2$ and this explain why in
$\dder{D}_1\cdot \dder{D}_2\cdot \dder{D}_0$ we have that $\dder{D}_1$ and $\dder{D}_2$ are no longer
independent.

@@

\begin{restatable}[Globality of independence]{lemma}{lemIndepGlobalLeft}
  \label{lem:indep-global-left}
  Let $\X$ be an $\mathcal{M}$-adhesive category and
  $(\X,\R)$ a strong enforcing DPO rewriting system.
  Consider a derivation
  $\der{D}=\{\dder{D}_i\}_{i=0}^2$ and suppose that $(i_0,i_1)$ is an independence pair between $\dder{D}_0$ and $\dder{D}_1$,
  $(a_0,a_1)$ one between $\dder{D}_1$ and $\dder{D}_2$ and
  $(e_0, e_1)$ one between $\dder{D}_0$ and
  $S_{a_0,a_1}(\dder{D}_2)$. Then the following properties hold true.
  \begin{enumerate}
  \item
    \label{lem:indep-global-left:1}
    $S_{e_0,e_1}(\dder{D}_0)$ and $S_{a_0,a_1}(\dder{D}_1)$ are
    switchable.

  \item
    \label{lem:indep-global-left:2}
    If $S_{i_0, i_1}(\dder{D}_0)\seqind \dder{D}_2$ with an independence pair $(\alpha_0, \alpha_1)$, then
    $S_{i_0,i_1}(\dder{D}_1)\seqind S_{\alpha_0, \alpha_1}(\dder{D}_2)$.
  \end{enumerate}
\end{restatable}

  %

@@

\section{Conclusions and further work}
\todo{VERY NICE CONCLUSIONS}

@@

\noindent \parbox{7.4cm}{
\begin{lemma}\label{lem:varie}Let $\mathcal{A}$ be a class of arrows stable under pullbacks, pushouts and containing all isomorphisms.  Suppose that, the left square below is $\mathcal{A}$-Van Kampen, while the vertical faces in the right cube are pullbacks. 	Supppose, moreover that $m\colon A\to C$ and $d\colon D'\to D$ are mono and that $d$ belongs to $\mathcal{A}$. Then $d\leq n$ if and only if $c \leq m$.
\end{lemma}}
	\parbox{6cm}{$\xymatrix@C=10pt@R=10pt{&&&&&A'\ar[dd]|\hole_(.65){a}\ar[rr]^{g'} \ar[dl]_{m'} && B' \ar[dd]^{b} \ar[dl]_{n'} \\ A \ar[dd]_{m}\ar[rr]^{g}&&B\ar[dd]^{n}&&C'  \ar[dd]_{c}\ar[rr]^(.7){f'} & & D' \ar[dd]_(.3){d}\\&&&&&A\ar[rr]|\hole^(.65){g} \ar[dl]^{m} && B \ar[dl]^{n} \\C\ar[rr]_{f} &&D&&C \ar[rr]_{f} & & D}$}

@@

	\smallskip \noindent
	$(\Leftarrow)$ Let $h:C\to A$ be such that $c=m\circ h$. By the $\mathcal{A}$-Van Kampen property the top face of the given cube is a pushout. Thus the dotted $k:D'\to B$ in the following diagram exists.
	\[\xymatrix@C=40pt{A' \ar@/_.5cm/[dd]_{a} \ar[d]^{m'} \ar[r]^{g'} & B' \ar[d]_{n'} \ar@/^.5cm/[dd]^{b} \\  C' \ar[d]^{h} \ar[r]_{f'} & D' \ar@{.>}[d]_{k}\\ A \ar[r]_g& B }\]
	Moreover, by construction we have
	\begin{align*}
		n\circ k \circ n' & = n\circ b=d\circ n'                                  \\
		n\circ k \circ f' & = n\circ g\circ h=f\circ m\circ h=f\circ c= d\circ f'
	\end{align*}
	We can therefore conclude that $n\circ k =d$.
\end{proof}}

@@

\full{ 
\begin{proof}
	We can begin noticing that $g$, being the pullback of $k$, is mono and in $\mathcal{A}$ too. Thus we can build the cube below, in which all the vertical faces are pullbacks, entailing that all the vertical arrows are in $\mathcal{A}$.
	\[\xymatrix@C=20pt@R=10pt{ & &X\ar[dl]_{f} \ar[ddd]^(.333333){\id{X}}|(.666666)\hole\ar[rrr]^{a} &&&A \ar[ddd]^{\id{A}} \ar[dl]^{h}\\& Y\ar[dl]_{\id{Y}} \ar[ddd]|(.333333)\hole_{\id{Y}} &&&B \ar[ddd]^{\id{B}} \ar[dl]_{\id{B}} \\Y \ar[ddd]_{g} \ar[rrr]^{b}&&& B\ar[ddd]^{k}\\&&X \ar[rrr]|(.34)\hole^{a}|(.67)\hole \ar[dl]_{f}&&& A \ar[dl]^{h}\\ & Y  \ar[dl]_{g}&&& B \ar[dl]^{k}\\ Z\ar[rrr]_{c} &&& C}\]
	By hypothesis the face is an $\mathcal{A}$-stable pushout and so its top one is a pushout. Using \Cref{lem:po1} we can conclude that the right half of the rectangle with which we have started is a pushout too.
\end{proof}}

@@

\begin{corollary}\label{prop:bal}
	If $\X$ is an $\mathcal{M}$-adhesive categories, then every epimorphisms in $\mathcal{M}$ is an isomorphisms. In particular, every adhesive category $\X$ is \emph{balanced}: if a morphism is monic and epic, then it is an isomorphism.
\end{corollary}

@@

\noindent
\parbox{11cm}{
	\begin{definition}[Pushout complement]
		Let $f\colon X\to Y$ and $g\colon Y\to Z$ be two composable arrows in a category $\X$. A \emph{pushout complement} for the pair $(f,g)$ is a pair $(h,k)$ with $h\colon X\to W$ and $k\colon W\to Z$ such that the square below commutes and it is a pushout.
	\end{definition}} \quad
\parbox{4cm}{\vspace{.1cm}
	$\xymatrix{X \ar[r]^{f} \ar[d]_{h}& Y \ar[d]^{g} \\ W \ar[r]_{k}& Z}$}\\

@@

Working in an $\mathcal{M}$-adhesive category we can amend the second defect.

@@

	Thus there exists a unique $t\colon A\to P$ as in the diagram
	\[\xymatrix{A \ar@{>->}[d]_{a} \ar@{.>}[dr]^{t}\ar[r]^{r} &B \ar@{>->}@/^.2cm/[dr]^{s} \\ M \ar@{>->}@/_.2cm/[dr]_{m}&P\ar[r]^{p_1}  \ar@{>->}[d]_{p_2}& N \ar@{>->}[d]^{n}\\ &X \ar[r]_{f} & Y}\]
	We can then consider the cube below, in which the front, right and back faces are pullbacks.
	\[\xymatrix@C=15pt@R=15pt{&A\ar@{>->}[dd]|\hole_(.65){a}\ar[rr]^{r} \ar[dl]_{t} && B \ar@{>->}[dd]^{b} \ar[dl]_{s} \\ P  \ar@{>->}[dd]_{p_2}\ar[rr]^(.7){p_1} & & N \ar@{>->}[dd]_(.3){n}\\&M\ar[rr]|\hole^(.65){h} \ar@{>->}[dl]^{m} && Q \ar@{>->}[dl]^{k} \\X \ar[rr]_{f} & & Y}\]
	We can notice that the left square is a pullback too. To see this it is enough to apply \Cref{lem:pb1} to the diagram
	\[\xymatrix{A  \ar@/^.4cm/[rr]^{s\circ r}\ar@{>->}[d]_{a} \ar[r]_{t}& \ar[r]_{p_1} P \ar@{>->}[d]^{p_2}& N \ar[d]^{c}\\ M \ar@/_.4cm/[rr]_{k\circ h}\ar@{>->}[r]^{m}& X \ar[r]^{f}& Y}\]

@@

\full{ 
\begin{proof}
	Consider the two pushout squares
	\[\xymatrix{M \ar[r]^{h_1}\ar@{>->}[d]_{m}& W_1 \ar@{>->}[d]^{k_1} & M \ar@{>->}[d]_{m} \ar[r]^{h_1}& W_2 \ar@{>->}[d]^{k_2}\\Y\ar[r]_{f}&Z &Y \ar[r]_{f}&Z}\]
	Since $\mathcal{M}$ is in $\mathcal{M}$, \cref{prop:pbpoad} guarantees that they are both pullbacks.
	Since $m\leq m$, by \Cref{lem:radj} we get that now entails that $k_1\leq k_2$ and $k_2\leq k_1$. Thus there exists an isomorphism $\phi\colon W_1\to W_2$ such that $k_1=k_2\circ \phi$. To see that $h_2=\phi\circ h_1$, we can compute:
	\[
		k_2\circ \phi \circ h_1  = k_1\circ h_1= n\circ m= k_2\circ h_2\]

@@

\label{propUnique-proof}
\begin{proof}
  Both pairs $(k, f)$ and $(k', f')$ are pushout complements of $l$
  and $n$, thus, \Cref{lem:pocomp} yields an isomorphism
  $\phi_D\colon D\to D'$. Computing we have
  \[
    g'\circ \phi_D \circ k= g' \circ k'=h'\circ r
  \]

@@

There is another
useful properties of weakly sequentially independent direct
derivations which is worth mentioning.

\noindent
\parbox{9.5cm}
{\begin{proposition}\label{lem:cose}Let
    $\der{D}=\{\dder{D}_i\}_{i=0}^1$ be a derivation, and suppose
    that $(i_0, i_1)$ is an independence pair between $\dder{D}_0$
    and $\der{D}_1$.  If $\der{E}=\{\dder{E}_i\}_{i=0}^1$ is a
    switch of $\der{D}$ along $(i_0, i_1)$, then there exists
    $q_0\colon P\to D_{\der{E},0}$ in the diagram
    below \end{proposition}}
\parbox{4cm}{$\xymatrix{D_{\der{E},0} \ar@{>->}[d]_{f_{\der{E},0}}&P
    \ar@{.>}[l]_{q_0} \ar[r]^{p_1} \ar@{>->}[d]_{p_0}& D_1
    \ar@{>->}[d]^{f_1}\\ G_{0}&D_0 \ar@{>->}[l]^{f_0}\ar[r]_{g_0}&
    G_1}$}

\begin{proof}
  Consider the arrow $u_1\colon K_1\to P$ obtained using
  \Cref{prop:tec}. It fits in the diagram:
\[		\xymatrix{K_1 \ar[r]^{u_1} \ar@{>->}[d]_{l_1}&P
	\ar[r]^{\id{P}}\ar@{>->}[d]_{p_0}& P\ar@{>->}[d]^{p_0}\\L_1
	\ar[d]_{\id{L_1}}\ar[r]^{i_1}&D_0 \ar[r]^{\id{D_0}}
	\ar[d]_{\id{D_0}}& D_{0} \ar@{>->}[d]^{f_0}\\
	L_1\ar[r]^{i_1} \ar@/_.4cm/[rr]_{m_{\der{E},0}}&D_0
	\ar@{>->}[r]^{f_0}&G_0}\]
